Realizability-Preserving DG-IMEX Method for the Two-Moment Model of Fermion Transport \tnoteref{support}

### 2018-04-20 Documents/M1/realizableFermionicM1/sections/imex.tex
Completed first pass in updating DGIMEXscheme.tex
Added more details for coefficients and c_sch.

To prove the positivity-preserving property of the IMEX scheme, Hu et al. \cite{hu_etal_2017} rewrite the stage values in Eq.~\eqref{imexStages} in the following form
\begin{equation}
  \vect{u}^{(i)}
  =\sum_{j=0}^{i-1}c_{ij}\Big[\,\vect{u}^{(j)}+\hat{c}_{ij}\,\dt\,\vect{\cT}(\vect{u}^{(j)})\,\Big]
  +a_{ii}\,\dt\,\f{1}{\tau}\,\vect{Q}(\vect{u}^{(i)}),\quad i=1,\ldots,s,
  \label{eq:imexStagesRewrite}
\end{equation}
where $c_{ij}$, and $\hat{c}_{ij}=\tilde{c}_{ij}/c_{ij}$ are computed from $\tilde{a}_{ij}$ and $a_{ij}$ (see \cite{hu_etal_2017} for details).  
(In Eq.~\eqref{eq:imexStagesRewrite}, $\vect{u}^{(0)}=\vect{u}^{n}$.)  
Two types of IMEX schemes are considered: \emph{type A} and \emph{type ARS}.  
For IMEX schemes of type A, the matrix $A$ is invertible.  
For IMEX schemes of type ARS, the first column of $A$ is zero and the submatrix consisting of the last $s-1$ rows and columns of $A$, $\hat{A}$, is invertible.  
For type A schemes, $\tilde{c}_{i0}=0$, and the positivity-preserving property follows from requiring $a_{ii}>0$ and $c_{i0}\ge0$ for $i=1,\ldots,s$, and $c_{ij},\tilde{c}_{ij}\ge0$, for $i=2,\ldots,s$, $j=1,\ldots,s-1$.  
For type ARS schemes, $c_{i1}=\tilde{c}_{i1}=0$, and positivity follows from requiring $a_{ii}>0$ and $c_{i0},\tilde{c}_{i0}\ge0$ for $i=1,\ldots,s$, and $c_{ij},\tilde{c}_{ij}\ge0$ for $i=3,\ldots,s$, $j=2,\ldots,i-1$.  
Such coefficients were given in \cite{hu_etal_2017} for GSA schemes of type A with $s=3$ and type ARS with $s=4$.  
(It was also proven that $s=3$ and $s=4$ are the necessary number of stages needed for GSA second-order positivity-preserving IMEX schemes of type A and type ARS, respectively.)

@@

We therefore proceed to design GSA positivity-preserving IMEX schemes, without the correction step, that perform better in the diffusion limit.  
To this end we define the vectors $\vec{\cJ}=(\cJ^{(1)},\ldots,\cJ^{(s)})^{T}$ and $\vec{\vect{\cH}}=(\vect{\cH}^{(1)},\ldots,\vect{\cH}^{(s)})^{T}$.  
In the context of IMEX schemes, the diffusion limit balance implies that the relation $A\,\vec{\vect{\cH}}=-\f{1}{3}\,\tau\,\tilde{A}\,\nabla\vec{\cJ}$ should hold to relate the stage values in Eq.~\eqref{imexStages}.  
For the final stage ($i=s$), 
\begin{equation}
  \vect{\cH}^{(s)}=-\f{1}{3}\,\tau\,\vect{e}_{s}^{T}A^{-1}\tilde{A}\vect{e}\,\nabla\cJ^{n}+\cO(\dt\,\tau),
\end{equation}
where $\vect{e}_{s}=(0,\ldots,0,1)^{T}$ and $\vect{e}=(1,\ldots,1)^{T}$.  
Thus, for $\vect{\cH}^{(s)}$ to be accurate in the diffusion limit, we require the IMEX coefficients to satisfy
\begin{equation}
  \vect{e}_{s}^{T}A^{-1}\tilde{A}\vect{e} = 1.
  \label{eq:diffusionCondition}
\end{equation}
Unfortunately, this requirement, together with the order conditions given by Eqs.~\eqref{orderConditions1} and \eqref{orderConditions2} (with $\alpha=0$), and the positivity conditions on $c_{ij}$ and $\tilde{c}_{ij}$, result in too many constraints.  
(\ee{Can we prove that it is impossible to find a second-order accurate, positivity-preserving IMEX scheme that is accurate in the diffusion limit?  Does the answer to this depend on $s$?})
(We are also concerned about increasing the number of stages, and thereby the number of implicit solves, since the implicit solve will dominate the computational cost of the IMEX scheme with more realistic collision operators.)
To reduce the number of constraints, and accommodate accuracy in the diffusion limit, we relax the requirement of overall second-order accuracy of the IMEX scheme.  
Instead, we only require the scheme to be second-order accurate in the streaming limit ($\vect{\cQ}=0$).  
This gives the order conditions
\begin{equation}
  \sum_{i=1}^{s}\tilde{w}_{i}=1
  \quad\text{and}\quad
  \sum_{i=1}^{s}\tilde{w}_{i}\,\tilde{c}_{i}=\f{1}{2}.  
  \label{eq:orderConditionsEx}
\end{equation}
We then seek to design GSA IMEX schemes with the following properties
\begin{itemize}
  \item Second-order in the streaming limit; i.e., satisfies Eq.~\eqref{eq:orderConditionsEx}.
  \item Well-behaved in the diffusion limit; i.e., satisfies Eq.~\eqref{eq:diffusionCondition}.
  \item Positivity-preserving (conditions given by \cite{hu_etal_2017}; with $c_{\mbox{\tiny Sch}}$ maximal).
  \item Few implicit stages.
\end{itemize}
Fortunately, IMEX schemes satisfying these properties are easy to find, and we provide an example of type ARS with $s=3$ (two implicit solves) in \ref{app:butcherTables} (scheme PARSD).  
In the streaming limit, this scheme is identical to the optimal second-order accurate strong-stability preserving Runge-Kutta method \cite{gottlieb_etal_2001}.  
It is also very similar to the scheme given in \cite{mcclareen_2008} (see scheme PC2 in \ref{app:butcherTables}), which is also a GSA IMEX scheme of type ARS with $s=3$.  
Scheme PC2 is second-order in the streaming limit, has been demonstrated to work well in the diffusion limit \cite{mcclareen_2008,radice_etal_2013}, and satisfies the positivity conditions, but $c_{\mbox{\tiny Sch}}=0$ (our primary motivation for finding an alternative).  
In Section~\ref{sec:numerical}, we show numerically that the accuracy of scheme PARSD is practically identical to the accuracy of scheme PC2.

>>>>

To prove the positivity-preserving property of the IMEX scheme, Hu et al. \cite{hu_etal_2017} rewrite the stage values in Eq.~\eqref{imexStages} in the following form
\begin{equation}
  \vect{u}^{(i)}
  =\sum_{j=0}^{i-1}c_{ij}\Big[\,\vect{u}^{(j)}+\hat{c}_{ij}\,\dt\,\vect{\cT}(\vect{u}^{(j)})\,\Big]
  +a_{ii}\,\dt\,\f{1}{\tau}\,\vect{Q}(\vect{u}^{(i)}),\quad i=1,\ldots,s,
  \label{eq:imexStagesRewrite}
\end{equation}
where $c_{ij}$, and $\hat{c}_{ij}=\tilde{c}_{ij}/c_{ij}$ are computed from $\tilde{a}_{ij}$ and $a_{ij}$.
(In Eq.~\eqref{eq:imexStagesRewrite}, $\vect{u}^{(0)}=\vect{u}^{n}$.)  
Two types of IMEX schemes are considered: \emph{type A} and \emph{type ARS}.  
For IMEX schemes of type A, the matrix $A$ is invertible.  
For IMEX schemes of type ARS, the matrix $A$ can be written as
\[\left( 
   \begin{matrix} 
       0 & 0 \\ 
       0 & \hat{A}
   \end{matrix}
\right)\]
with $\hat{A}$ be invertible.
For $i=1,\ldots,s$, the coefficients
\begin{align*}
  c_{i0}
  =&\left\{
  \begin{array}{cl}
    1 - \sum_{j=1}^{i-1}c_{ij}  & \text{for type $A$,} \\
    1 - \sum_{j=2}^{i-1}c_{ij}  & \text{for type $ARS$;}
  \end{array}
  \right.\\
  \tilde{c}_{i0}
  =&\left\{
  \begin{array}{cl}
    0  & \text{for type $A$,} \\
   \tilde{a}_{i1} + \sum_{l=2}^{i-1}a_{il}\tilde{b}_{l1}  & \text{for type $ARS$;}
  \end{array}
  \right.\\
  c_{i1}
    =&\left\{
    \begin{array}{cl}
      \sum_{l =1}^{i-1} a_{il}b_{l1} & \text{for type $A$,} \\
      0  & \text{for type $ARS$;}
    \end{array}
    \right.\\
    \tilde{c}_{i1}
    =&\left\{
      \begin{array}{cl}
       \tilde{a}_{i1} + \sum_{l=2}^{i-1}a_{il}\tilde{b}_{l1} & \text{for type $A$,} \\
       0  & \text{for type $ARS$;}
      \end{array}
      \right.
\end{align*}
and the rest coefficients were defined as in \cite{hu_etal_2017}.
For type A schemes, the positivity-preserving property follows from requiring $a_{ii}>0$ and $c_{i0}\ge0$ for $i=1,\ldots,s$, and $c_{ij},\tilde{c}_{ij}\ge0$, for $i=2,\ldots,s$, $j=1,\ldots,s-1$.  
For type ARS schemes, the positivity follows from requiring $a_{ii}>0$ and $c_{i0},\tilde{c}_{i0}\ge0$ for $i=2,\ldots,s$, and $c_{ij},\tilde{c}_{ij}\ge0$ for $i=3,\ldots,s$, $j=2,\ldots,i-1$.  
Such coefficients were given in \cite{hu_etal_2017} for GSA schemes of type A with $s=3$ and type ARS with $s=4$.  
(It was also proven that $s=3$ and $s=4$ are the necessary number of stages needed for GSA second-order positivity-preserving IMEX schemes of type A and type ARS, respectively.)

@@

We therefore proceed to design GSA positivity-preserving IMEX schemes, without the correction step, that perform better in the diffusion limit.  
To this end we define the vectors $\vec{\cJ}=(\cJ^{(1)},\ldots,\cJ^{(s)})^{T}$ and $\vec{\vect{\cH}}=(\vect{\cH}^{(1)},\ldots,\vect{\cH}^{(s)})^{T}$.  
In the context of IMEX schemes, the diffusion limit balance implies that the relation $A\,\vec{\vect{\cH}}=-\f{1}{3}\,\tau\,\tilde{A}\,\nabla\vec{\cJ}$ should hold to relate the stage values in Eq.~\eqref{imexStages}.  
For the final stage ($i=s$), 
\begin{equation}
  \vect{\cH}^{(s)}=-\f{1}{3}\,\tau\,\vect{e}_{s}^{T}A^{-1}\tilde{A}\vect{e}\,\nabla\cJ^{n}+\cO(\dt\,\tau),
\end{equation}
where $\vect{e}_{s}=(0,\ldots,0,1)^{T}$ and $\vect{e}=(1,\ldots,1)^{T}$.  
Thus, for $\vect{\cH}^{(s)}$ to be accurate in the diffusion limit, we require the IMEX coefficients to satisfy
\begin{equation}
  \vect{e}_{s}^{T}A^{-1}\tilde{A}\vect{e} = 1.
  \label{eq:diffusionCondition}
\end{equation}
Unfortunately, this requirement, together with the order conditions given by Eqs.~\eqref{orderConditions1} and \eqref{orderConditions2} (with $\alpha=0$), and the positivity conditions on $c_{ij}$ and $\tilde{c}_{ij}$, result in too many constraints.  
(\ee{Can we prove that it is impossible to find a second-order accurate, positivity-preserving IMEX scheme that is accurate in the diffusion limit?  Does the answer to this depend on $s$?})
(We are also concerned about increasing the number of stages, and thereby the number of implicit solves, since the implicit solve will dominate the computational cost of the IMEX scheme with more realistic collision operators.)
To reduce the number of constraints, and accommodate accuracy in the diffusion limit, we relax the requirement of overall second-order accuracy of the IMEX scheme.  
Instead, we only require the scheme to be second-order accurate in the streaming limit ($\vect{\cQ}=0$).  
This gives the order conditions
\begin{equation}
  \sum_{i=1}^{s}\tilde{w}_{i}=1
  \quad\text{and}\quad
  \sum_{i=1}^{s}\tilde{w}_{i}\,\tilde{c}_{i}=\f{1}{2}.  
  \label{eq:orderConditionsEx}
\end{equation}
We then seek to design GSA IMEX schemes with the following properties
\begin{itemize}
  \item Second-order in the streaming limit; i.e., satisfies Eq.~\eqref{eq:orderConditionsEx}.
  \item Well-behaved in the diffusion limit; i.e., satisfies Eq.~\eqref{eq:diffusionCondition}.
  \item Positivity-preserving (conditions given by \cite{hu_etal_2017}; with $c_{\mbox{\tiny Sch}}$ maximal).
  \item Few ($\le3$) implicit stages.
\end{itemize}
Fortunately, IMEX schemes satisfying these properties are easy to find, and we provide an example of type ARS with $s=3$ (two implicit solves) in \ref{app:butcherTables} (scheme PARSD).  
In the streaming limit, this scheme is identical to the optimal second-order accurate strong-stability preserving Runge-Kutta method \cite{gottlieb_etal_2001}.  
It is also very similar to the scheme given in \cite{mcclareen_2008} (see scheme PC2 in \ref{app:butcherTables}), which is also a GSA IMEX scheme of type ARS with $s=3$.  
Scheme PC2 is second-order in the streaming limit, has been demonstrated to work well in the diffusion limit \cite{mcclareen_2008,radice_etal_2013}, and satisfies the positivity conditions, but $c_{\mbox{\tiny Sch}}=0$ (our primary motivation for finding an alternative).  
In Section~\ref{sec:numerical}, we show numerically that the accuracy of scheme PARSD is practically identical to the accuracy of scheme PC2.
